---
title: Import estimates from multiple NONMEM runs
author: Kyle Baron
date: 10/02/2022
description: > 
  You can use $NMEXT and $NMXML to import estimates from multiple NONMEM runs
  into a single mrgsolve model.
categories: 
- pkpd
---

The `$NMEXT` and `$NMXML` blocks allow you to import estimates (THETA, SIGMA, and OMEGA)
from a NONMEM run. So if you are running a population PK model in NONMEM and translating 
to mrgsolve, you should never need to copy the final estimate but rather read them in 
using these blocks. 

If you are running sequential PK/PD or parent/metabolite models, you might need
to pull estimates from _multiple_ NONMEM runs. This is easy to do with multiple
`$NMEXT` (or `$NMXML`) blocks. 

## Import the first (PK) model

First, grab the estimates from the PK model; let's call it run 101. The `$NMEXT`
block could look like this


```{r, eval = FALSE}
$NMEXT
run = 101
project = "../model/nonmem"
root = "cppfile"
```

This will find the file `201.ext` in the `../model` directory which is
taken to be relative to the location of the mrgsolve model file (using the
`root` argument). If the paths are a little confusing, you can try installing
the [here](https://CRAN.R-project.org/package=here) package and using

```{r, eval = FALSE}
$NMEXT
run = 101
project = here::here("model/nonmem")
```

This would require an Rstudio project file in the (project) root directory 
and then we would locate the NONMEM run relative to that project root
(e.g. in `<project-root>/model/nonmem`). 

Once the first model is imported, we now have access to `THETA1`, `THETA2`
etc from the PK model. 

## Import the second (PD) model


Now, we import estimates from the PD model run (let's call it 201) using another 
call to `$NMEXT`


```{r, eval = FALSE}
$NMEXT
run = 201
project = here::here("model/nonmem")
tname = "theta"
```

This is the second import in the same mrgsolve model file, so we change the 
handle for referencing the `THETA` estimates: now the PD model uses `theta1`, 
`theta2` etc so that we can retain the THETA numbering for both the PK and PD 
models, remembering that the PD model uses `theta1` (lower case) while the 
PK model uses `THETA1` (upper case). Note that we _have_ to change the handle
(to something like `theta`) otherwise mrgsolve will throw an error because
there would be two `THETA1` in the parameter list.


